---
title: "Formative Assessment 2"
author: "Ramilo, Zion John Yousef"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(wbstats)
```

Import
```{r}
who <- who2
population <- wb_data(indicator = "SP.POP.TOTL")
population$country <- replace(population$country, population$country == "United States", "United States of America")
```

Shapes of the datasets
```{r}
cat("TB WHO Data:\n")
cat("Rows: ",dim(who)[1],"\nColumns: ",dim(who)[2])
cat("\n\nPopulation Data:\n")
cat("Rows: ",dim(population)[1],"\nColumns: ",dim(population)[2])
```
Variable Typesof Population
```{r}
sapply(population, class)
colSums(is.na(population))
```
Fix any anomalies
```{r}
population2 <- subset(population, select = -c(unit,obs_status,footnote,iso2c,iso3c,last_updated)) %>% 
  na.omit()
glimpse(population2)
colSums(is.na(population2))
```
Identify variables (WHO dataset)
```{r}
colnames(who)
who2 <- pivot_longer(who,cols=sp_m_014:rel_f_65,names_to = "new_ep_f014", values_to = "cases") %>% 
  mutate(cases = replace_na(cases, 0))
glimpse(who2)
```
Seperate values
```{r}
who_tidy <- who2 %>% 
  separate(new_ep_f014,into=c("tb_type","sex","age_range"), sep="_")
glimpse(who_tidy)
```

Identify variables (Population dataset)
```{r}
colnames(population2)
```
Perform pivot operation on population data
```{r}
population3 <- population2 %>% 
  pivot_wider(names_from = date,values_from = SP.POP.TOTL)
glimpse(population3)
```
Population Tidy Ver.
```{r}
population_tidy <- population2 %>% 
  rename(
    population=SP.POP.TOTL,
    year = date
  )
glimpse(population_tidy)
```
Joining Datasets
```{r}
tuberculosis <- who_tidy %>%
  group_by(country, year, sex, tb_type, age_range) %>%
  summarise(cases = sum(cases, na.rm = TRUE), .groups = "drop")

tuberculosis <- merge(tuberculosis, population_tidy, by = c("country", "year"), all.x = TRUE)
glimpse(tuberculosis)
```

United States TB Cases by Sex
```{r}
us_tb_cases<- tuberculosis %>% 
  filter(country == "United States of America" & year >= 2001) %>% 
  group_by(sex) %>% 
  summarise(sum_cases = sum(cases))
glimpse(us_tb_cases)
```
Males are more prone to having TB within the United States during this 21st Century.

Aggregate cases into cases by 100K
```{r}
tuberculosis2 <- tuberculosis %>% 
  group_by(year, sex, tb_type, age_range) %>% 
  mutate(cases_per_100k = (cases / population) * 100000)
glimpse(tuberculosis2)
```
Identify Highest and Lowest Cases by 100k by country and year
```{r}
highest_cases_per_100k <- tuberculosis2 %>% 
  group_by(country,year) %>% 
  summarise(sum_cases_per_100k = sum(cases_per_100k), .groups = "drop") %>% 
  arrange(desc(sum_cases_per_100k))
```
Top 10 highest:
```{r}
head(highest_cases_per_100k,10)
```
Top 10 Lowest
```{r}
tail(highest_cases_per_100k%>%na.omit(),10)  
```
Plot the total cases per 100k as a function of year for China, India, and the United States.
```{r}
cases_per_100k_CIUS <- highest_cases_per_100k %>% 
  filter(country == "China" | country == "India" | country == "United States of America") %>% 
  arrange(year)
ggplot(cases_per_100k_CIUS,aes(x=year,y=sum_cases_per_100k,color=country))+
  geom_line()+
  scale_y_log10()
```
During the 21st century TB was seen high growth as compared to 2000s and below implying that there was a point within the 21st century that enabled TB to increase within the contires.

Compare distributions of total cases per 100k (summed over years, sexes, and TB types) across age groups
```{r}
tuberculosis_ageRange <- tuberculosis2 %>%
  group_by(age_range) %>%
  summarise(total_cases_per_100k = sum(cases_per_100k, na.rm = TRUE), .groups = "drop")
ggplot(tuberculosis_ageRange, aes(x = total_cases_per_100k)) +
  geom_density(alpha = 0.5) +
  scale_y_log10() + 
  labs(
    title = "Distribution of TB Cases per 100k by Age Group",
    x = "Total Cases per 100k ",
    y = "Density"
  ) +
  theme_minimal()
```
The graph shows a stable constant number of TB cases per 100k.

Create a plot to evaluate whether the number of cases per 100k in 2000 was related to a country’s population
```{r}
tb_population <- tuberculosis2 %>% 
  filter(year==2000) %>% 
  group_by(population) %>% 
  summarise(total_cases_per_100k = sum(cases_per_100k, na.rm = TRUE), .groups = "drop")

ggplot(tb_population,aes(x=population,y=total_cases_per_100k))+
  geom_line()+ 
  scale_x_log10()
```
The plot shows very high volatility as population grows the more unpredictable the total cases per 100k is produced.
